---
title: "Week 8 Tutorial"
subtitle: "Applied Analytical Statistics 2025/26"
author: "Paul Röttger and Mikhail Korneev"
output:
  html_document:
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this tutorial, we explore practical applications of **panel data and fixed effects**, as well as issues related to **autocorrelation and time dependence**.

We use data from the American Community Survey (ACS) covering the years 2014–2024. The full dataset is available at: https://data.census.gov/. For all questions regarding the tutorial materials, please contact Mikhail Korneev at mikhail.korneev@reuben.ox.ac.uk. 

By the end of this tutorial, you will be able to:

- Construct **panel datasets**
- Estimate regression models with **fixed effects**
- Detect and address **time dependence**
- Model and correct for **autocorrelation** with AR(1)

### Panel datasets and Fixed Effects

#### Intuition: Simulated Data

Panel datasets are defined by their structure. Consider a data simulation below that models the relationship between firms adopting a certain policy (e.g., diversity in hiring) and their profitability. 

What makes it a panel dataset? 

```{r panel data simulation, message=FALSE} 

# Simulated panel dataset for firm profitability

# Set seed for replication
set.seed(123)

# packages
library(dplyr)

# parameters
N <- 100; T <- 20
years <- 1:T
shock_years <- c(5, 11, 17)
shock_mags <- c(-0.25, -0.4, -0.3)

# firm covariates
firms <- tibble(
  firm = paste0("F", 1:N),
  region = sample(c("NA","EU","AS","OT"), N, replace = TRUE, prob = c(.4,.3,.2,.1)),
  industry = sample(c("Software","Hardware","Services","Security"), N, replace = TRUE, prob = c(.5,.2,.2,.1)),
  prod = rnorm(N),      # productivity 
  gov  = rnorm(N),      # governance quality 
  exp  = rnorm(N)       # founder experience
)

# firms adopt policy X with probability p 
# p depends only on observed covariates (staggered adoption)
firms <- firms %>%
  mutate(p_adopt = plogis(0.4*gov + 0.3*prod - 0.2),
         ever = rbinom(N, 1, p_adopt),
         adopt_year = ifelse(ever == 1, sample(6:15, N, replace = TRUE), NA_integer_))

# build panel
panel <- expand.grid(firm = firms$firm, year = years, KEEP.OUT.ATTRS = FALSE) %>%
  as_tibble() %>%
  left_join(firms, by = "firm") %>%
  mutate(policy = ifelse(!is.na(adopt_year) & year >= adopt_year, 1, 0))

# one-year shocks (same for all firms in that year), outcome, factors
panel <- panel %>%
  mutate(
    shock = case_when(
      year == shock_years[1] ~ shock_mags[1],
      year == shock_years[2] ~ shock_mags[2],
      year == shock_years[3] ~ shock_mags[3],
      TRUE ~ 0
    ),
    eps = rnorm(n(), 0, 0.2),
    # firm fixed effect is exact linear function of observed covariates (no unobserved noise)
    profitability = 1 + 0.5*prod + 0.3*gov + 0.2*exp + shock + 0.12*policy + eps,
    yearf = factor(year),
    firmf = factor(firm)
  )

```

```
The simulated data is a panel dataset because it contains repeated observations for the same cross-sectional units (firms) over multiple time periods (years). Specifically, the dataset follows 100 firms (N = 100) observed annually over 20 years (T = 20).

This is also a balanced panel, since it contains observations for all units and time periods. 

```

In the simulated panel dataset, which variables vary over time but are common across firms? Which variables vary across firms but remain constant over time? 

```
The only variable that changes over time and is the same for all firms is the shock variable. These shocks occur in specific years (years 5, 11, and 17) and affect every firm equally in those years. This represents a common macroeconomic shock.

Several characteristics differ between firms but remain constant over time for each firm. These include: region, industry, prod (productivity), gov (governance quality), exp (founder experience), ever (whether the firm ever adopts the policy), adopt_year (year of adoption, once assigned). These characteristics are fixed attributes of each firm and do not change across years in the simulation. 

```
When analyzing the relationship between policy adoption and profitability, which variables are confounders and why? 

```
Given the data simulation, governance and productivity are the key confounders, since they affect the probability of policy adoption and profitability. 

Conversly, experience is not a confounder because it is not correlated with the probability of policy adoption. 

```

The panel data structure allows us to control for confounders by controlling for time-invariant characteristics and unit-invariant effects. When some confounders are unobserved, two-way fixed effects (TWFE) model outperforms multivariate OLS by addressing the omitted variable bias.  

To see this in practice, implement two **multivariate pooled OLS models** that measure the association between policy adoption and profitability: 

- For model 1, control for **all confounders** but do not include the fixed effects. 
- For model 2, assume that governance is unobserved and exclude it from the model. 


```{r OLS, warning=FALSE, message = FALSE} 
library(stargazer)
library(sandwich)
library(lmtest)

# Multivariate pooled OLS with observed covariates 
model_OLS_1 <- lm(profitability ~ policy + gov + exp + 
                 prod + region + industry + shock,
               data = panel)

# Multivariate pooled OLS, but governance is not observed 
model_OLS_2 <- lm(profitability ~ policy + exp + 
                 prod + region + industry + shock,
               data = panel)

# cluster-robust vcov
vcov_cluster_OLS_1 <- vcovCL(model_OLS_1, cluster = panel$firm)
vcov_cluster_OLS_2 <- vcovCL(model_OLS_2, cluster = panel$firm)

# extract clustered SE for stargazer
se_cluster_OLS_1 <- sqrt(diag(vcov_cluster_OLS_1))
se_cluster_OLS_2 <- sqrt(diag(vcov_cluster_OLS_2))

# display model output 
stargazer(model_OLS_1, model_OLS_2,
          type = "text",
          keep = "policy",
          se = list(se_cluster_OLS_1, se_cluster_OLS_2),
          column.labels = c("Model 1", "Model 2"),
          dep.var.labels = "Profitability", 
          add.lines = list(
            c("Controls", "Yes", "Partial"),
            c("Firm Fixed Effects", "No", "No"),
            c("Year Fixed Effects", "No", "No")
          )) 

```
What is the interpretation of the coefficient? Which model best captures the true known (by design) association between the policy and profitability and why? 

```
According to model 1 with all controls, a single unit increase in policy (i.e., adopting a policy) is associated with a 0.121 unit increase in profitability. The association is significant at the 1% level. This essentially captures the true known association of 0.12 that was used in the simulation. 

When governance is not included in the regression equation, the coefficient is biased upwards due to the OVB. A single unit increase in policy is associated with a 0.23 units increase in profitability. The coefficient is significant at the 1% level, but substantially exceeds the true known simulated relationship. 

```

Pooled OLS with known confounders (model 1) uses within and between variation. What does this mean in practice? How is this reflected in our example? 

```
Within variation: Changes over time within the same firm. Between variation: Differences across firms.

Between-firm variation comes from time-invariant firm characteristics such as prod, gov, exp, region, and industry. These differ across firms but do not change over time. When we include these variables in pooled OLS, the model compares more productive firms to less productive firms.

Within-firm variation comes from variables that change over time, such as policy (after adoption) and shock (macro shocks in certain years). When estimating the policy effect, pooled OLS partly relies on how profitability changes within a firm once it adopts the policy.

```

Now, implement a **two-way fixed effects model**. To do that, regress profitability on the policy, plus year and firm dummies. Do not include any other confounders. 

Remember to **cluster standard errors** at firm level. 

```{r TWFE model, warning=FALSE}

# Two-way FE using dummy variables (equivalent to firm + year FE)
model_FE <- lm(profitability ~ policy + factor(firm) + factor(year),
               data = panel)

# cluster-robust vcov
vcov_cluster_FE <- vcovCL(model_FE, cluster = panel$firm)

# extract clustered SE for stargazer
se_cluster_FE <- sqrt(diag(vcov_cluster_FE))

stargazer(model_OLS_1, model_OLS_2, model_FE,
          type = "text",
          keep = "policy",
          
          # use clustered SEs
          se = list(se_cluster_OLS_1, se_cluster_OLS_2,  se_cluster_FE),
          
          # edit the table 
          column.labels = c("OLS (model 1)", "OLS (model 2)", "Two-way FE"),
          dep.var.labels = "Profitability",
          add.lines = list(
            c("Controls", "Yes", "Partial", "No"),
            c("Firm Fixed Effects", "No", "No", "Yes"),
            c("Year Fixed Effects", "No", "No", "Yes")
          ))

```

Compare the outcome of the three models. Do the coefficients differ? Why yes or why not? 

```
Both the OLS specification with all controls in column 1 and the two-way FE model in column 3 capture the true known relationship between policy adoption and profitability. 

This similarity arises because the fixed effects model controls for all time-invariant firm characteristics (via firm fixed effects) and for common shocks affecting all firms in a given year (via year fixed effects). In OLS model 1, firm-level characteristics such as productivity, governance, and experience are observed and included in the pooled specification, and the time shocks are explicitly modeled.

Notably, the two-way FE model outperforms the OLS specification that treats governance as an unobserved charactertistic and omits it from the specification; the coefficient is closer to the known association (less bias) and the standard errors are smaller (greater precision). Therefore, fixed effects can help us get unbiased coefficients even when some data on time-invariant characteristics is not available. 

```

What (by-design) assumptions did we make about the simulated data? Would the results of the FE model be similar if these assumptions were violated? 

```
We made several strong by-design assumptions in constructing the simulated data. The fixed effects (FE) estimator performs well in this setting precisely because these assumptions hold. If they were violated, the FE results could be biased or inconsistent.

The central assumption is strict exogeneity. That is, we assumed that the error term is uncorrelated with the regressors in all time periods. This implies:

- No reverse causality (profitability does not influence policy adoption),
- No anticipation effects (firms do not adopt based on expected future shocks),
- No omitted time-varying confounders.

```

In our simulation, firms are observed over 20 years. In reality, what theoretical assumptions might be violated? 

```
Unit FEs control for time-invariant heterogeneity. Over 20 years, we expect that few things are time-invariant. For instance, in practice, firm governance changes due to leadership changes, productivity changes due to technological investments. 

Furthermore, macroeconomic conditions, regulatory shifts, and technological disruptions are likely to the associated with profitability changes. As such, a more realistic model should pay greater attention to time-varying confounders. 

```

#### Practice: ACS Database 

To provide a practical example, we now focus on the ACS database that collects information on an extensive list of socio-economic characteristics of the US labour force. 

We selected several variables for the tutorial to explore **economic determinants of self-employment**. Go ahead and download the data from GitHub.

What makes this a **panel dataset**? What are the units of observation? Is this a balanced panel? State the number of units and time periods. 

```{r ACS data} 

library(readr)

df_ACS <- read_csv(
  "https://raw.githubusercontent.com/paul-rottger/aas-2026-public/refs/heads/main/week8/tutorial/week8_tutorial_sample.csv"
)

```

```
The sample is a panel dataset because it contains repeated observations for the same entities across multiple time periods. In the case of the ACS data, the units of observation are 51 U.S. states, the time dimension is year, and states are observed for 9 years between 2014 and 2024 (excluding 2020). Each state is observed in multiple years, allowing us to track changes within states over time.

Because we have observations for all variables for each state in each available year, the dataset constitutes a balanced panel. The only exception is that the year 2020 is missing due to COVID-19 disruptions. Aside from this omitted year, the panel is balanced across states and time.

```

In our sample, what could be the **time-invariant** characteristics? What could be the **unit-invariant** characteristics? 

```
Time-invariant characteristics may include features such as a state's geographic location, long-run market structure, underlying demographic composition, institutional environment, or persistent legal frameworks. These characteristics are relatively stable and unlikely to change substantially over the ten-year period from 2014 to 2024.

Entity-invariant characteristics (time-specific shocks that affect all states equally) may include nationwide economic shocks, federal legislation, monetary policy changes, international crises, or other macroeconomic events that influence the entire country in a given year.

```

Self-employment trends differ between states in the US. Use ggplot to visualise self-employment for the five largest states: California, Texas, Florida, New York, and Pennsylvania.  

What might be the story about the time trend? 

```{r self-employment trend}
library(ggplot2)

# filter by state name
df_5_states <- df_ACS %>%
  filter(geo %in% c("California",
                    "Texas",
                    "Florida",
                    "New York",
                    "Pennsylvania"))
# plot the trends
ggplot(df_5_states, aes(x = year, y = pct_self_employed, color = geo, group = geo)) +
  geom_line(size = 1) +
  geom_point(size = 1.5, alpha = 0.8) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 8)) +
  labs(
    title = "Self-employment rate over time — five largest states",
    subtitle = "California, Texas, Florida, New York, Pennsylvania",
    x = "Year",
    y = "Pct self-employed",
    color = "State"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold")
  )

```

```
All five states display persistently different levels of self-employment between 2014 and 2024, indicating the importance of controlling for time-invariant state characteristics using state fixed effects. California consistently had the highest self-employment rate, likely reflecting structural economic factors, while Pennsylvania remained comparatively low throughout the period.

Self-employment increased notably in Texas, Florida, and New York between 2019 and 2021, possibly due to the economic disruption caused by COVID-19. This common shift highlights the need for year fixed effects to account for shocks that affect all states but vary over time.

```

What parameter may be associated with self-employment? Select one predictor and implement a two-way fixed effects model. 

```{r TWFE ACS, warning=FALSE}
library(stargazer)

# TWFE
ACS_1 <- lm(pct_self_employed ~ I(est_median_hh_income/1000)  # express in thousands to simplify the interpretation
            + factor(year) + factor(geo),
            data = df_ACS)

# cluster-robust vcov
vcov_cluster_ACS_1 <- vcovCL(ACS_1, cluster = df_ACS$geo)

# extract clustered SE for stargazer
se_cluster_ACS_1 <- sqrt(diag(vcov_cluster_ACS_1))

# display model output 
stargazer(ACS_1,
          type = "text",
          se = list(se_cluster_ACS_1),
          keep = "est_median_hh_income",
          column.labels = c("TWFE"),
          dep.var.labels = "pct_self_employed", 
          add.lines = list(
            c("Controls", "No"),
            c("Firm Fixed Effects", "Yes"),
            c("Year Fixed Effects", "Yes")
          ))

```

What is the interpretation of the coefficient? 

```
A $1,000 increase in median household income within a state — relative to that state’s own long-run average and relative to the national average in that year — is associated with a 0.043 percentage point decrease in that state’s self-employment rate, holding constant all time-invariant differences across states and common nationwide year shocks.

```

Note that the TWFE only captures entity- and time-invariant characteristics. If we have reasons to believe that some confounders are time-variant, leaving them out of the model may introduce the Omitted Variable Bias (OVB). 

Improve the model by adding 1-2 controls that might be associated with both self-employment and your independent variable, and that vary across time. 

```{r TWFE ACS with controls, warning=FALSE}

# TWFE with controls
ACS_2 <- lm(pct_self_employed ~ I(est_median_hh_income/1000) # express in thousands of dollars
              + pct_labor_force + pct_work_from_home + pct_poverty_all +
              factor(year) + factor(geo),
            data = df_ACS)

# cluster-robust vcov
vcov_cluster_ACS_2 <- vcovCL(ACS_2, cluster = df_ACS$geo)

# extract clustered SE for stargazer
se_cluster_ACS_2 <- sqrt(diag(vcov_cluster_ACS_2))

# Display model output
stargazer(ACS_1, ACS_2,
          type = "text",
          keep = c("est_median_hh_income", "pct_labor_force", "pct_work_from_home"),
          se = list(se_cluster_ACS_1, se_cluster_ACS_2),
          column.labels = c("TWFE", "TWFE with controls"),
          dep.var.labels = "pct_self_employed", 
          add.lines = list(
            c("Controls", "No", "Yes"),
            c("Firm Fixed Effects", "Yes", "Yes"),
            c("Year Fixed Effects", "Yes", "Yes")
          ))

```

Does the size of the coefficient change? Why? 

```
The marginal change between columns (1) and (2) suggests that the association between median household income and self-employment is largely robust to the inclusion of labor force participation and work-from-home rates, indicating that neither variable meaningfully confounds the baseline relationship.

Note that our TWFE models produce very high R-squared. This is largely driven by the inclusion of state and year fixed effects, which absorb substantial variation in the outcome across units and over time.

```

What percentage of self-employment does the model predict for California in 2022 if your predictors are kept at national averages for that year? 

```{r TWFE prediction} 
# prediction using the ACS_2 model 
pred <- ACS_2$coefficients[1] +                                   # intercept
        ACS_2$coefficients["factor(geo)California"] +             # California FE
        ACS_2$coefficients["factor(year)2022"] +                  # 2022 FE
        ACS_2$coefficients["I(est_median_hh_income/1000)"] * mean(df_ACS$pct_private_workers[df_ACS$year == 2022], na.rm = TRUE) +
        ACS_2$coefficients["pct_labor_force"]      * mean(df_ACS$pct_unemployed[df_ACS$year == 2022], na.rm = TRUE) +
        ACS_2$coefficients["pct_work_from_home"]     * mean(df_ACS$pct_gov_workers[df_ACS$year == 2022], na.rm = TRUE)

pred

```

### Autocorrelation and Time Dependence

Working with observations across multiple time periods might introduce a new problem: **autocorrelation**. 

Returning to the simulated data and the ACS examples, do you think autocorrelation might be a concern? 

```
By construction, the dependent variable in the simulated data is not autocorrelated. However, in real life, profitability might be time dependent because the underlying drivers (brand strength, operational efficiency, market position) change slowly. 

For the ACS data, it is likely that the percentage of self-employed individuals is correlated over time. Self-enployment is likley to be a persistent feature state-level economies and is constrained by imperfect labour market mobility. 

```

Diagnosing autocorrelation requires a sufficiently large sample (usually T > 50-100). For simplicity, we will therefore switch to a new example. 

Run the code below to simulate a dataset on the association between economic growth and unemployment. 

```{r data time series} 

# Set seed for reproducibility
set.seed(123)

# time series
T <- 120  # 10 years of monthly data
dates <- seq.Date(from = as.Date("2015-01-01"),
                  by = "month",
                  length.out = T)

# economic growth with autocorrelation 
growth <- numeric(T)
growth[1] <- rnorm(1, 0.2, 0.5)  # draws 1 value from N(0.2, 0.25)

for (t in 2:T) {
  growth[t] <- 0.6 * growth[t-1] + rnorm(1, 0, 0.4)
}

# unemployment as a function of economic growth and past unemployment
unemployment <- numeric(T)
unemployment[1] <- 6 + rnorm(1, 0, 0.3)

for (t in 2:T) {
  unemployment[t] <- 0.8 * unemployment[t-1] - 
                     0.3 * growth[t] + 
                     rnorm(1, 0, 0.2)
}

# dataframe
df_ts <- tibble(
  date = dates,
  economic_growth = growth,
  unemployment = unemployment
)

```

#### Biased coefficients and SEs

Autocorrelation occurs when the **error terms in a regression model are correlated across observations**. Among other factors, this can result from omitting time-varying variables and misspecifying functional form and dynamics.

In our simulation, lagged coefficients are included in the output formula. This is typical for inert processes, such as macroeconomic trends. 

By itself, omitting lagged coefficients produces **incorrect functional form** and leads to **biased estimates**. To see this, implement a naive OLS regression model that measures the association between economic growth and unemployment. 

```{r naive OLS}

# naive model 
OLS_TS <- lm(unemployment ~ economic_growth, df_ts)
summary(OLS_TS)

```

Does the model correctly capture the known association between unemployment and economic growth? 

```
According to the naive OLS, a single unit increase in economic growth produces a 0.17 decrease in unemployment. This is substantially different from the known by design association. 

```

More importantely for statistical inference, omitted lagged variables create *autocorrelation in the error term**. This **biases the standard errors**. Because autocorrelation is left out of the regression equation, OLS residuals are time-dependent. Plot residuals from the naive OLS model. 

```{r naive OLS residuals}

df_ts$residuals <- resid(OLS_TS)

library(ggplot2)

ggplot(df_ts, aes(x = date, y = residuals)) +
  geom_line(color = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals from OLS Model",
       x = "Date",
       y = "Residuals") +
  theme_minimal()

```

#### Diagnosing Autocorrelation 

To diagnose autocorrelation, we can use Autocorrelation Function (ACF) and Partial Autocorrelation Function (PACF) plots that measure correlation between a series and its lags, and formal tests. 

First, use the ACF and the PACF to check for serial correlation in residuals. What is the interpretation of the plots? 

```{r ACF and PACF} 
par(mfrow = c(1, 2))

acf(df_ts$residuals,
    main = "ACF of OLS Residuals")

pacf(df_ts$residuals,
    main = "PACF of OLS Residuals")

```

```
The ACF graph reveals a clear serial correlation. There is a large positive autocorrelation at lag 1 (~0.6), and the autocorrelation then gradually decays over several lags.Several early lags also exceed the confidence bands.

This slow decay pattern is characteristic of an AR(1) process (or low-order AR process). It suggests the residuals are not white noise.

The PACF indicates that there is a single strong spike at lag 1. After lag 1, most partial autocorrelations fall inside the confidence bands. This is indicative of an AR(1) process. 

```

Alternatively, we can use the Durbin-Watson test for time dependence. Run the test and interpret the results.  

```{r DW}

library(lmtest)
dwtest(OLS_TS)

```

```
The p-valye of the Durbin-Watson test is below 0.05. We can therefore reject the null hypothesis that the true autocorrelation in the OLS model is 0. 

```

#### AR(1) Model 

Time-series analysis is a complex field of econometrics and statistical inference. In practice, there is no single solution to autocorrelation, and the right model depends on the underlying assumptions about the data and the relationship. 

For our example, a simple OLS with a lagged dependent variable will perform relatively well to get unbiased estimates. Because it uses a single lag of the dependent variable, this model is known as the **AR(1) model**.  

Make a lag of the dependent variable and re-run the OLS specification. What is the interpretation of the coefficients? 

```{r AR(1)} 

# make lag
df_ts <- df_ts %>% arrange(date)
df_ts$unemp_lag1 <- dplyr::lag(df_ts$unemployment, 1)
df_ts$economic_growth_lag1 <- dplyr::lag(df_ts$economic_growth, 1)

# drop first obs
df_ar <- df_ts %>% filter(!is.na(unemp_lag1))

# estimate
model_ar <- lm(unemployment ~ economic_growth + unemp_lag1, data = df_ar)
summary(model_ar)

# For robust SEs 
library(lmtest)
library(sandwich)

coeftest(model_ar, vcov = NeweyWest(model_ar, lag = 4))

```

```
A one-unit increase in economic growth is associated with a 0.25 decrease in unemployment, holding past unemployment constant. This is close to the true simulated effect of −0.3 and is highly statistically significant (p < 0.001).

A one-unit increase in unemployment last period is associated with a 0.776 increase in current unemployment, holding growth constant. This confirms strong serial dependence and is close to the true known simulation parameter of 0.8.

```

Construct an ACF graph to check for autocorrelation in the residuals for the new model. What is the interpretation of the new ACF graph? 

```{r ACF AR(1)}

# Residual ACF
resid_ar <- resid(model_ar)
acf(resid_ar)

```

```
By including lagged terms in the AR(1) model, we remove time dependence in the residuals. 

```

